---
title: "Student Debt Across the USA and how does New College of Florida Students Compare Over Time (2019–2025)"
author: "Jorge Urbaez"
output:
  html_document:
    theme: simplex
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: kable
editor_options:
  markdown:
    wrap: 72
---

```{r, warning=FALSE, message=FALSE}
# Load packages here: 
library(tidyverse)
library(ggplot2)
library(mosaic)
library(readr)
library(dplyr)

# Import the data here:

files <- list.files("data", pattern = "\\.csv$", full.names = T)


C_2019 <- read_csv(files[1],na = c("-999", "not reported", "", "PS" ) )
C_2020 <- read_csv(files[2],na = c("-999", "not reported", "", "PS" ) )
C_2024 <- read_csv(files[3],na = c("-999", "not reported", "", "PS" ) )
C_2025 <- read_csv(files[4],na = c("-999", "not reported", "", "PS" ) )


C_2019 <- C_2019|>
  mutate(
    REGION = case_when(
      REGION == "U.S.S.S" ~ 0, REGION == "U.S" ~ 0,
      REGION == "NE" ~ 1,      REGION == "ME" ~ 2,
      REGION == "GL" ~ 3,      REGION == "Plains" ~ 4,
      REGION == "SE" ~ 5,      REGION == "SW" ~ 6,
      REGION == "Rockies" ~ 7, REGION == "West" ~ 8,
      REGION == "Outlying" ~ 9,
      TRUE ~ NA_real_
    ),
    Year = 2019
  )

C_2024 <- C_2024|>
  mutate(
    REGION = case_when(
      REGION == "U.S.S.S" ~ 0, REGION == "U.S" ~ 0,
      REGION == "NE" ~ 1,      REGION == "ME" ~ 2,
      REGION == "GL" ~ 3,      REGION == "Plains" ~ 4,
      REGION == "SE" ~ 5,      REGION == "SW" ~ 6,
      REGION == "Rockies" ~ 7, REGION == "West" ~ 8,
      REGION == "Outlying" ~ 9,
      TRUE ~ NA_real_
    ),
    Year = 2024
  )
  

C_2025 <- C_2025|>
            transform(
              ADM_RATE = as.numeric(ADM_RATE), 
              UGDS = as.numeric(UGDS),
              COSTT4_A = as.numeric(COSTT4_A), 
              MN_EARN_WNE_P10 = as.numeric(MN_EARN_WNE_P10),
              MD_EARN_WNE_P10 = as.numeric(MD_EARN_WNE_P10), 
              MN_EARN_WNE_MALE0_P10 = as.numeric(MN_EARN_WNE_MALE0_P10),
              MN_EARN_WNE_MALE1_P10 = as.numeric(MN_EARN_WNE_MALE1_P10), 
              MD_EARN_WNE_P6 = as.numeric(MD_EARN_WNE_P6),
              MN_EARN_WNE_P6 = as.numeric(MN_EARN_WNE_P6), 
              MN_EARN_WNE_MALE0_P6 = as.numeric(MN_EARN_WNE_MALE0_P6),
              MN_EARN_WNE_MALE1_P6 = as.numeric(MN_EARN_WNE_MALE1_P6), 
              GRAD_DEBT_MDN_SUPP = as.numeric(GRAD_DEBT_MDN_SUPP),
              RPY_3YR_RT_SUPP = as.numeric(RPY_3YR_RT_SUPP), 
              LO_INC_RPY_3YR_RT_SUPP = as.numeric(LO_INC_RPY_3YR_RT_SUPP),
              MD_INC_RPY_3YR_RT_SUPP = as.numeric(MD_INC_RPY_3YR_RT_SUPP),
              HI_INC_RPY_3YR_RT_SUPP = as.numeric(HI_INC_RPY_3YR_RT_SUPP), 
              C150_4_POOLED_SUPP = as.numeric(C150_4_POOLED_SUPP),
              C200_4_POOLED_SUPP = as.numeric(C200_4_POOLED_SUPP),
              RET_FT4_POOLED_SUPP = as.numeric(RET_FT4_POOLED_SUPP), 
              TUITIONFEE_OUT = as.numeric(TUITIONFEE_OUT),
              TUITIONFEE_IN  = as.numeric(TUITIONFEE_IN)
                      )|>
  mutate(isNCF = grepl("New College of Florida", INSTNM))
 

year <- '2025'
current_year <- C_2025
```

## Introduction

This is a rough report looking at college costs and debt. The purpose is
to explore the College Scorecard data and see what patterns we can find
for the `r year`

We are especially interested in three questions:

1.  What does the distribution of median student debt look like across
    all U.S. institutions?\n
2.  How do costs of attendance vary across regions of the country?\n
3.  How does New College of Florida compare to other schools in terms of
    cost, debt, and repayment, and how has this changed over time?

Some of the results will focus on the year `r year`, but we might also
want to check previous years (2019) to see how things have changed.

### Data Preparation

The data we are working with comes from the College Scorecard, a public
dataset of U.S. colleges. We are using versions from 2019 on up through
the current year `r year`. Each file has information on cost of
attendance, debt, repayment rates, and other metrics. You can access the
data here: you can access here: <https://collegescorecard.ed.gov/data/>.

There are a few issues to note in this data set. For example, some
variables use special codes instead of leaving values blank. The
variable **DEBT_MDN_SUPP**, which is the median debt per institution,
uses the value -999 to indicate missing data. The variable **COSTT4_A**,
which is the annual total cost of attendance, uses the phrase "not
reported" to indicate missing data, which means it may not load as
numeric. In more recent years (2025 and after), the Scorecard sometimes
shows values as `"PS"`, meaning *privacy suppressed*. This happens when
the sample size is too small and the Department of Education does not
release the true value.

We also need to think about the region variable. Right now, the regions
are given as abbreviations, but originally they were coded as numbers.
The table below shows the numeric codes, the abbreviations, and the
states in each region.

```{r}
 regions <- tribble(
  ~Code, ~Region, ~`States`,
  0, "U.S.S.S", "U.S. Service Schools",
  1, "NE", "CT, ME, MA, NH, RI, VT",
  2, "ME", "DE, DC, MD, NJ, NY, PA",
  3, "GL", "IL, IN, MI, OH, WI",
  4, "Plains", "IA, KS, MN, MO, NE, ND, SD",
  5, "SE", "AL, AR, FL, GA, KY, LA, MS, NC, SC, TN, VA, WV",
  6, "SW", "AZ, NM, OK, TX",
  7, "Rockies", "CO, ID, MT, UT, WY",
  8, "West", "AK, CA, HI, NV, OR, WA",
  9, "Outlying", "AS, FM, GU, MH, MP, PR, PW, VI"
)

```

`r regions`

We will use the region variable in several sections of the report.

### Results

The results section is broken into 4 sub-sections. First, we look at
overall median debt across all schools. Then we break the analysis of
debt down by region, take a closer look at students' abilities to repay
their debts, and finally we examine at how NCF compares to other
schools.


## Overall median debt across all schools in `r year`

```{r}
current_year|> 
  summarise(`Median Debt in (USD)` = median(DEBT_MDN_SUPP, na.rm = TRUE))
```

### Median Student Debt by Region

```{r}
current_year|> 
  group_by(REGION)|>
  summarise(`Median Student Debt (USD)` = median(DEBT_MDN_SUPP, na.rm = TRUE))|>
  arrange(desc(`Median Student Debt (USD)`))
```

### Ability to repay debt

```{r}
current_year |> 
  summarise(`Median Debt Repayment Rate` = median(RPY_3YR_RT_SUPP, na.rm = TRUE))

```

### Ability of NCF students to repay their debt compared to students from other institutions

```{r}
#New College of Florida
ncf_debt_repay <- current_year |> 
  filter(isNCF == TRUE) |> 
  summarise(
    debt = median(DEBT_MDN_SUPP, na.rm = TRUE),
    repay = median(RPY_3YR_RT_SUPP, na.rm = TRUE)
  )

# All other schools
all_schools_debt_repay <- current_year |> 
  filter(isNCF == FALSE) |> 
  summarise(
    debt = median(DEBT_MDN_SUPP, na.rm = TRUE),
    repay = median(RPY_3YR_RT_SUPP, na.rm = TRUE)
  )
```

New college of Florida student median debt in `r year` was
`r format(round(ncf_debt_repay$debt, 0), scientific = FALSE)`, a
`r format(round(ncf_debt_repay$debt - all_schools_debt_repay$debt, 3),  scientific = FALSE)` (USD)
difference in comparison to other school's debt, and New College of
Florida students debt repayment rate after 3 years was around
`r format(round(ncf_debt_repay$repay, 3),  scientific = FALSE)` percentage points, `r format(round(ncf_debt_repay$repay - all_schools_debt_repay$repay, 3),  scientific = FALSE)` percentage points difference compared to Students from other institutions.

One important question when looking at colleges is: how much debt do
students graduate with? Median student debt is one of the best ways to
measure this, since it is less sensitive to extreme outliers than the
mean. This question matters because student debt has long-term
consequences for graduates and their ability to repay loans, buy homes,
or continue their education.

We first look at the overall distribution of median debt across all
schools in the data set. Then we will highlight the median debt for New
College of Florida (NCF) for comparison.

```{r warning=FALSE}
ncf_debt <- current_year |> 
  filter(isNCF == TRUE) |> 
  summarise(median_debt = median(DEBT_MDN_SUPP, na.rm = TRUE))


ggplot(current_year, aes(x=DEBT_MDN_SUPP, y="" ))+
  
  geom_boxplot(fill= "lightblue", 
               color = "black")+
  geom_point(
    data = filter(current_year, isNCF == TRUE), 
    aes(x = DEBT_MDN_SUPP),
    color = "red"
  )+
  geom_text(
    data = filter(current_year, isNCF == TRUE),
    aes(x = DEBT_MDN_SUPP, y = "" , label = "New College of Florida"),
    color = "red",
    vjust = 1.5
  )+
  labs(
    title = "Distribution of Median Student Debt Across All Schools", 
    y= "",
    x = "Median Student Debt (USD)"
    )+
  theme_classic()


```

The median student debt for NCF in `r year` was `r ncf_debt[[1]]` (USD).

### Numerical summary and histogram

```{r warning=FALSE}

# summary and histogram of DEBT_MDN_SUPP goes here

current_year|>
  summarise(
    Min = min(DEBT_MDN_SUPP, na.rm = TRUE),
    Q1 = quantile(DEBT_MDN_SUPP, 0.25, na.rm = TRUE),
    Median = median(DEBT_MDN_SUPP, na.rm = TRUE),
    Q3 = quantile(DEBT_MDN_SUPP, 0.75, na.rm = TRUE),
    Max = max(DEBT_MDN_SUPP, na.rm = TRUE),
    Mean = mean(DEBT_MDN_SUPP, na.rm = TRUE),
    SD = sd(DEBT_MDN_SUPP, na.rm = TRUE),
    Total = sum(length(DEBT_MDN_SUPP)),
    Valid = sum(!is.na(DEBT_MDN_SUPP)),
    Missing = sum(is.na(DEBT_MDN_SUPP))
  )

ggplot(current_year, aes(x = DEBT_MDN_SUPP))+
  geom_histogram(fill = "lightblue", color = "black", bins = 22)+
  labs(
    title = "Distribution of Median Student Debt Across All Schools",
    x = " Median Student Debt (USD) ",
    y = "Number of Schools",
  )+
  theme_classic()

```

## Cost by Region in `r year`

Another key question is how the cost of attending college varies
depending on the region of the country. This is important because
students often choose schools close to home, and costs can differ widely
across regions. By comparing costs across regions we can see whether
some areas consistently have more expensive schools than others. It's
also useful to consider the sample size and amount of missing data in
each region.

```{r}
ncf_COSTT4_A <- current_year|>
  filter(isNCF == TRUE)|>
  select(COSTT4_A)
```

In the `r year` data set, the cost for NCF was `r format(round(ncf_COSTT4_A[[1]] , 3) ,  scientific = FALSE)` (USD).
This will be used as a reference line in the plot below.

### Boxplot of cost by region

The graph below shows side-by-side boxplots of the total cost of
attendance for each region. The red line marks NCF's cost for
comparison.

```{r warning=FALSE}
# boxplot of COSTT4_A by REGION with a red line for NCF’s cost
# i decided to turn back into string category, because i think that will make easier for readers to see the comparison without having to add a key to the graph or people having to go up the report it also includes other sections where i prefer to use string key instead of numbers

reg <- current_year|>
  mutate(
    REGION = case_when(
      REGION == 0 ~ "U.S.S.S",
      REGION == 1 ~ "NE",
      REGION == 2 ~ "ME",
      REGION == 3 ~ "GL",
      REGION == 4 ~ "Plains",
      REGION == 5 ~ "SE",
      REGION == 6 ~ "SW",
      REGION == 7 ~ "Rockies",
      REGION == 8 ~ "West",
      REGION == 9 ~ "Outlying",
      TRUE ~ NA_character_
    )
  )

ggplot(reg, aes(x = REGION, y = COSTT4_A))+
  geom_boxplot(fill= 'lightblue')+
  geom_hline(yintercept = ncf_COSTT4_A[[1]], color = "red")+
  labs(
    title = "Cost of Attendance by Region",
    x = "Region" ,
    y = "Cost of Attendance (USD)"
  )+
  theme_classic()
```

### Summary of highest and lowest cost schools

As we can see from the above summaries, the SouthEast (where NCF is
located) had some of the highest costing colleges, with outliers far
above the rest of the institutions. However, there's also a great deal
of variability, with several low-cost colleges as well. It's difficult
to see from the graph exactly what the most and least expensive colleges
are, so we'll pull this information out into a separate table.

The schools with the overall highest and lowest cost in `r year` were:

```{r}
# table showing INSTNM, REGION, COSTT4_A for the min and max schools

COSTT4_A_min_max <- reg|>
  dplyr::select(
    Institution = INSTNM, 
    Region = REGION,
    `Cost of Attendance` = COSTT4_A
    )|>
  filter(
    `Cost of Attendance`== min(`Cost of Attendance`, na.rm = T) | (`Cost of Attendance` == max(`Cost of Attendance`, na.rm = T))
    )
```
`r COSTT4_A_min_max`

## Comparing Metrics by Region `r year`

It is often useful to compare multiple financial measures side by side,
to see whether patterns hold across different variables. In particular,
we might ask whether regions that have higher costs of attendance also
tend to have higher levels of debt. Here we focus on two regions with
very different types of institutions: the Southeast (where NCF is
located) and New England.

This question is interesting because students choosing between schools
in these regions may want to know whether they are likely to graduate
with similar levels of debt, or whether cost and debt behave differently
depending on the region.

From a coding perspective, this section was more challenging, because
the data had to be restructured so that both cost and debt could be
displayed on the same plot.

### Side-by-side boxplots of cost and debt in Southeast vs New England

```{r warning=FALSE}
# create 4 boxplots on the same graph:
# Median debt and cost of attendance for Southeast and New England regions
reg_long <- reg |>
  filter(REGION %in% c("SE", "NE")) |>
  select(
    Region = REGION, 
    `Median Debt` = DEBT_MDN_SUPP, 
    `Cost of Attendance` = COSTT4_A) |>
  pivot_longer(
    cols = c(`Median Debt`, `Cost of Attendance`),
    names_to = "Var_name",
    values_to = "Value"
  )

reg_long|>
  ggplot(aes(x = Region, y = Value), fill = Region)+
  geom_boxplot()+
  facet_wrap(~Var_name)+
  labs(
    title = "Median Debt and Cost of Attendance in Southeast vs New England",
    x = "Region",
    y = "USD",
    fill = "Region"
  )+
  theme_classic()

```

This concludes the report on college costs and debt.
